---
title: "Heatmap Visualization sheet"
date: "2025-03-25"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # Show code (set to FALSE to hide)
  message = FALSE,    # Hide package loading messages
  warning = FALSE,    # Hide warnings
  error = TRUE        # allow knitting to continue if errors occur
)
# setwd("J:/Dropbox/Prague/Teaching/Course genomics of speciation/2024-2025/RNA-Seq practical/Heatmap visualization")
```

Would you like to do a heatmap? Why?

Which input data would you use for heatmap and Why? Should you use all genes or only a subset?

# Load required libraries

```{r}
# Install required packages if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
if (!requireNamespace("RColorBrewer", quietly = TRUE)) install.packages("RColorBrewer")

# Load libraries
library(pheatmap)      # for creating heatmaps
library(RColorBrewer)  # for color palettes

# Check your current directory
getwd()

# Set your working directory
setwd("J:/Dropbox/Prague/Teaching/Course genomics of speciation/2024-2025/RNA-Seq practical/Heatmap visualization")
```

# Import data

```{r}
# Load expression data
tpm_matrix <- read.csv("your_read_counts.csv", row.names = 1)

# (Optional) Filter genes for better visualization — e.g., top variable genes
top_var_genes <- apply(tpm_matrix, 1, var)
top_genes <- names(sort(top_var_genes, decreasing = TRUE))[1:50]  # top 50 variable genes
mat_log <- log2(tpm_matrix[top_genes, ] + 1)  # log-transform for better contrast
```

# Create a basic Heatmap

```{r}
pheatmap(mat_log,
         scale = "row",                     # normalize expression across each gene
         clustering_method = "ward.D2",     # method for hierarchical clustering
         show_rownames = TRUE,              # show gene names
         show_colnames = TRUE,              # show sample names
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(50),
         main = "Heatmap of Top Variable Genes")
```

# (Optional) Add column annotations by grouping onto the Heatmap

Again, 'grouping' column in the metadata represents the experimental condition you want to use to check if samples show similar expression profile or not.

```{r}
# Load sample metadata
metadata_subset <- read.csv("your_sample_metadata.csv", row.names = 1)

# Create annotation data frame for column coloring
annotation_df <- data.frame(grouping = metadata_subset$grouping)
rownames(annotation_df) <- rownames(metadata_subset)  # sample IDs should match column names in mat_log

# Heatmap with sample annotations
pheatmap(mat_log,
         scale = "row",
         annotation_col = annotation_df,
         clustering_method = "ward.D2",
         show_rownames = TRUE,
         show_colnames = TRUE,
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(50),
         main = "Heatmap with grouping annotation")
```

Do you observe any noticeable differences in gene expression patterns between sample groupings?
